---
title: "compare"
author: "Gina"
format: html
editor: visual
---

```{r, echo = F, messages = F}
library(tidyverse)
```

# Introduction to irrigation energy

There may
be two sources of water:

1.  surface

2.  ground

Ground water will have a \'head\' associated with the depth of the well. Both will have \'head\' associated
with the pump pressure.  

The main components defining the irrigation energy use are:

1.  The pump pressure and, if ground water, the well depth

2.  The energy source for the pumping/moving of water (diesel,
    electricity, solar, etc.)

3.  The amount of irrigation water applied

# Scenario

-   A flood irrigated field

-   Diesel powered pump

-   200 foot well

-   25 PSI pump

-   \$1 energy unit (to make back-calculating the amount
    of fuel easy)

-   Corn for grain

-   1 acre

-   20 ac-in applied per acre `0`

-   The zipcode is used to auto-fill some NRCS values which we overwrite, so it shouldn\'t
    matter what value you put in. However I put in 81428.

    ```{r}
    well_depth <- 200
    pump_press <- 25
    water_applied <- 20
    ```

# NRCS tool

The NRCS has a tool that will estimate the energy used for irrigation: <https://ipat.sc.egov.usda.gov/Default.aspx>

When using the above inputs, the estimated energy cost is \$67. I take this to mean I will have used 67 gallons of diesel:

*picture is not pasting*

According to Table 1 of FTM, 1 gallon of diesel contains 138,490 BTUs/gal.

```{r}
nrcs_btu <- 67 * 138490
print(nrcs_btu)
```

This irrigation scenario requires `r format(nrcs_btu, big.mark=",", scientific=FALSE)` BTUs/ac**.** Just for ballpark estimates, Table 1 of FTM says there are 22.7 lbs of co2e per gallon of diesel (although this is from combustion of the fuel, only).

```{r}
nrcs_co2 <- 67 * 22.7
print(nrcs_co2)
```

So this irrigation scenario releases `r format(nrcs_co2, big.mark = ",", scientific = F)` lbs co2e/ac.

# FTM equations

The total head, $H$ is expressed as the sum of the head from the pump pressure and the pumping depth. To calculate the head we therefore just need to do some unit conversions.

First, convert the pump pressure to meters:

```{r}
# 2.31 ft head / 1 psi
# 0.305 m head / 1 ft head
# pump is 25 psi
print(press_head <- pump_press  * 2.31 * 0.305)

```

Next, convert the pump depth to meters:

```{r}
# depth  is 200 ft
print(depth_head <- 200 * 0.305)
```

Sum them together :

```{r}
print(head <- press_head + depth_head)

```

This is then used in the pumping energy equation:

$$energy = \frac{H*C_{pump}*(acre-inches_{applied})*25.4 \frac{mm}{in} * (number-of-acres-in-field) * (0.404 \frac{ha}{acre})}{E_{pump}*E_{irrigation}*E_{gearhead}*E_{power-unit}}$$


Where:

$$H = head\\
C_{pump} = 0.0979\\
E_{pump} = 0.75\\
E_{irrigation} = 1\\
E_{gearhead} = 0.95\\
E_{power-unit} = 1$$

The units are very strange, and I cannot follow what is being canceled. Apparently this produces a value in MJs per field? In our case, since we have a 1 acre field:

*Get constants in to R*

```{r}
# constants
c_pump <- 0.0979
e_pump <- 0.75
e_irr <- 1
e_gear <- 0.95
e_power <- 1

mm_per_in <- 25.4
m_per_in <- mm_per_in/100
ha_per_ac <- 0.404

```

*Calculate energy*

```{r}

print(ftm_energy_mj <- (head * c_pump * water_applied * mm_per_in * 1 * 0.404) / (e_pump * e_irr*e_gear*e_power))

```

There are 948 BTUs in 1 MJ

```{r}
print(ftm_btu <- ftm_energy_mj * 948)
```

According to FTM equations, this irrigation scenario requires `r format(ftm_btu, big.mark=",", scientific=FALSE)` BTUs/ac**.**

If we instead use the meters to inches conversion:

```{r}
print(ftm_btu2 <- (head * c_pump * water_applied * m_per_in * 1 * 0.404) / (e_pump * e_irr*e_gear*e_power)*948)
```

As a summary:

```{r}

tibble(method = c("nrcs", "FTM-mm", "FTM-m"),
       energy_BTUs_ac = c(nrcs_btu, ftm_btu, ftm_btu2)) %>% 
  mutate_if(is.numeric, round, 0)
```
